---
title: "Assessment 2"
output: html_document
date: "2024-09-13"
---

# Assessment 2 

Write a 3-page report on the distribution of tower sites discussing the strengths and weakness of the current tower representation. Please create 2 visualizations. You are welcome to use any additional data.


Weakness: continental/hemispherical representation.
  Figure: 
  
Strength:  
  Figure:  


Load libraries

```{r}

library(sf)
library(AOI)
library(ggplot2)
library(tidyverse)
library(lwgeom)

```


Load data

```{r}

FluxNet <- read.csv('FluxNet_Sites_2024.csv')

```


### Plot FLuxNet towers by hemisphere (N&S)

```{r}

# Make an object with AOI polygons for all terrestrial land area, as in workshop 2
aoi.terrestrial <- aoi_get(country= c("Europe","Asia" ,"North America", "South America", "Australia","Africa", "New Zealand")) %>%
st_transform( '+init=epsg:4326' )

# Remove rows where LOCATION_LAT or LOCATION_LONG are empty strings or NA
FluxNet <- FluxNet[!(FluxNet$LOCATION_LAT == "" | 
                     FluxNet$LOCATION_LONG == "" | 
                     is.na(FluxNet$LOCATION_LAT) | 
                     is.na(FluxNet$LOCATION_LONG)), ]

# Add column identifying northern and southern hemisphere towers
FluxNet$Hemisphere <- as.factor(ifelse(FluxNet$LOCATION_LAT > 0, "Northern", "Southern"))

# Count of number of towers in each hemisphere
hemisphere_counts <- table(FluxNet$Hemisphere)
n_northern <- hemisphere_counts["Northern"]
n_southern <- hemisphere_counts["Southern"]

# Convert the data frame to an sf object
FluxNetTowers <- st_as_sf(FluxNet,
                          coords = c("LOCATION_LONG", "LOCATION_LAT"),
                          crs = "+init=epsg:4326")

# Plot the terrestrial AOI polygons and FLuxNet towers by hemisphere
ggplot() +
  geom_sf(data = aoi.terrestrial) +
  geom_sf(data = FluxNetTowers,
          aes(color = Hemisphere),
          size = 1) +
  scale_color_manual(values = c("blue", "red")) +
  theme_minimal() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", size = 0.5) +
  annotate("text", x = -140, y = 10, label = paste("n = ", n_northern), size = 5, color = "blue", fontface = "bold") +
  annotate("text", x = -140, y = -10, label = paste("n =", n_southern), size = 5, color = "red", fontface = "bold") +
  ggtitle("FluxNet Tower Distribution by Hemisphere") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```

### Calculate land area by hemisphere

```{r}



```

