---
title: "Assessment 4 Bonus: Climate Comparison of Acadia and Zion National Parks"
output: html_document
date: "2024-09-25"
---

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(terra)
library(remotes)
library(tidyverse)
library(AOI)
library(climateR)
library(sf)
library(tidyterra)
library(ggplot2)
library(viridis)
library(ggpubr)
library(ggspatial)
library(exactextractr)
library(patchwork)

```

## Get Boundary polygons for ANP and ZNP

```{r message=FALSE, warning=FALSE}
# Read the NPS boundary shapefile
nps <- st_read('nps_boundary.shp')

# Extract Acadia National Park (ANP) polygon
ANP <- nps[nps$UNIT_NAME == 'Acadia National Park', ]
ANP <- st_transform(ANP, crs = "EPSG:4326")

# Extract Zion National Park (ZNP) polygon
ZNP <- nps[nps$UNIT_NAME == 'Zion National Park', ]
ZNP <- st_transform(ZNP, crs = "EPSG:4326")

```


## Download Climate Nornals for ANP and ZNP

and calculate total annual precipitation, mean maximum temperature, and mean minimum temperature for each park. 

```{r message=FALSE, warning=FALSE}
# Download climate normals for ANP
ANP_normals <- getTerraClimNormals(AOI = ANP, varname = c("ppt", "tmax", "tmin"))

# Calculate Total Annual Precipitation
ANP_TAP <- sum(ANP_normals$ppt)

# Calculate Mean Annual Maximum Temperature
ANP_mean_tmax <- mean(ANP_normals$tmax)

# Calculate Mean Annual Minimum Temperature
ANP_mean_tmin <- mean(ANP_normals$tmin)

# Reproject raster data to match ANP CRS
ANP_TAP <- project(ANP_TAP, "EPSG:4326")
ANP_mean_tmax <- project(ANP_mean_tmax, "EPSG:4326")
ANP_mean_tmin <- project(ANP_mean_tmin, "EPSG:4326")

###############################################################################

# Download climate normals for ZNP
ZNP_normals <- getTerraClimNormals(AOI = ZNP, varname = c("ppt", "tmax", "tmin"))

# Calculate Total Annual Precipitation
ZNP_TAP <- sum(ZNP_normals$ppt)

# Calculate Mean Annual Maximum Temperature
ZNP_mean_tmax <- mean(ZNP_normals$tmax)

# Calculate Mean Annual Minimum Temperature
ZNP_mean_tmin <- mean(ZNP_normals$tmin)

# Reproject raster data to match ZNP CRS
ZNP_TAP <- project(ZNP_TAP, "EPSG:4326")
ZNP_mean_tmax <- project(ZNP_mean_tmax, "EPSG:4326")
ZNP_mean_tmin <- project(ZNP_mean_tmin, "EPSG:4326")

```


## Update: Interpoaltion and Resampling

Use bilinear interpolation to resample the climate rasters at a finer resolution.

This aligns the raster data more precisely with the park boundary polygons for more accurate masking and analysis. 

```{r message=FALSE, warning=FALSE}

# Resample and Mask for ANP

# Define a finer resolution (adjust as needed)
finer_res <- 0.001  # Degrees; adjust based on your computational capacity

# Create a new raster template with the finer resolution
template_ANP <- rast(
  xmin = xmin(ANP_TAP),
  xmax = xmax(ANP_TAP),
  ymin = ymin(ANP_TAP),
  ymax = ymax(ANP_TAP),
  resolution = finer_res,
  crs = crs(ANP_TAP)
)

# Resample the raster data
ANP_TAP_resampled <- resample(ANP_TAP, template_ANP, method = "bilinear")
ANP_tmax_resampled <- resample(ANP_mean_tmax, template_ANP, method = "bilinear")
ANP_tmin_resampled <- resample(ANP_mean_tmin, template_ANP, method = "bilinear")

# Mask the resampled raster data with the ANP polygon
exact_ANP_TAP <- mask(ANP_TAP_resampled, vect(ANP))
exact_ANP_tmax <- mask(ANP_tmax_resampled, vect(ANP))
exact_ANP_tmin <- mask(ANP_tmin_resampled, vect(ANP))

################################################################################

# Resample and Mask for ZNP

# Create a new raster template with the finer resolution
template_ZNP <- rast(
  xmin = xmin(ZNP_TAP),
  xmax = xmax(ZNP_TAP),
  ymin = ymin(ZNP_TAP),
  ymax = ymax(ZNP_TAP),
  resolution = finer_res,
  crs = crs(ZNP_TAP)
)

# Resample the raster data
ZNP_TAP_resampled <- resample(ZNP_TAP, template_ZNP, method = "bilinear")
ZNP_tmax_resampled <- resample(ZNP_mean_tmax, template_ZNP, method = "bilinear")
ZNP_tmin_resampled <- resample(ZNP_mean_tmin, template_ZNP, method = "bilinear")

# Mask the resampled raster data with the ZNP polygon
exact_ZNP_TAP <- mask(ZNP_TAP_resampled, vect(ZNP))
exact_ZNP_tmax <- mask(ZNP_tmax_resampled, vect(ZNP))
exact_ZNP_tmin <- mask(ZNP_tmin_resampled, vect(ZNP))

```

## Update: 
### a different (bettter?) way to extract raster values over polygons

The exact_extract() function of the exactextractr package in R is designed to extract raster values over polygons, accounting for the exact area of overlap between raster cells and polygons. 

```{r message=FALSE, warning=FALSE}

# Calculate mean values within ANP using exactextractr
ANP_TAP_exact_mean <- exact_extract(ANP_TAP, ANP, 'mean')
ANP_tmax_exact_mean <- exact_extract(ANP_mean_tmax, ANP, 'mean')
ANP_tmin_exact_mean <- exact_extract(ANP_mean_tmin, ANP, 'mean')

# Calculate mean values within ZNP using exactextractr
ZNP_TAP_exact_mean <- exact_extract(ZNP_TAP, ZNP, 'mean')
ZNP_tmax_exact_mean <- exact_extract(ZNP_mean_tmax, ZNP, 'mean')
ZNP_tmin_exact_mean <- exact_extract(ZNP_mean_tmin, ZNP, 'mean')

```

## Display Calculated Values

```{r message=FALSE, warning=FALSE}

# Display the results for ANP
cat("Acadia National Park (ANP):\n")
cat("Mean annual precipitation:", round(ANP_TAP_exact_mean, 2), "mm\n")
cat("Mean annual maximum temperature:", round(ANP_tmax_exact_mean, 2), "°C\n")
cat("Mean annual minimum temperature:", round(ANP_tmin_exact_mean, 2), "°C\n\n")

# Display the results for ZNP
cat("Zion National Park (ZNP):\n")
cat("Mean annual precipitation:", round(ZNP_TAP_exact_mean, 2), "mm\n")
cat("Mean annual maximum temperature:", round(ZNP_tmax_exact_mean, 2), "°C\n")
cat("Mean annual minimum temperature:", round(ZNP_tmin_exact_mean, 2), "°C\n")

```

## Plot Climate Data and Park Boundaries

```{r fig.width=15, fig.height=15, message=FALSE, warning=FALSE}

# Common theme adjustments
common_theme <- theme_minimal() +
  theme(
    legend.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

# Visualization for ANP
# Precipitation
ANP_precip_plot <- ggplot() +
  geom_spatraster(data = exact_ANP_TAP) +
  geom_sf(data = ANP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Acadia National Park: Total Annual Precipitation", fill = "Precipitation (mm)") +
  coord_sf(expand = FALSE) +
  common_theme

# Maximum Temperature
ANP_tmax_plot <- ggplot() +
  geom_spatraster(data = exact_ANP_tmax) +
  geom_sf(data = ANP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Acadia National Park: Mean Annual Max Temperature", fill = "Temperature (°C)") +
  coord_sf(expand = FALSE) +
  common_theme

# Minimum Temperature
ANP_tmin_plot <- ggplot() +
  geom_spatraster(data = exact_ANP_tmin) +
  geom_sf(data = ANP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Acadia National Park: Mean Annual Min Temperature", fill = "Temperature (°C)") +
  coord_sf(expand = FALSE) +
  common_theme

# Arrange ANP plots

ANP_Climate <- ggarrange(ANP_precip_plot, ANP_tmax_plot, ANP_tmin_plot, ncol = 1)

##############################################################################

# Visualization for ZNP
# Precipitation
ZNP_precip_plot <- ggplot() +
  geom_spatraster(data = exact_ZNP_TAP) +
  geom_sf(data = ZNP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Zion National Park: Total Annual Precipitation", fill = "Precipitation (mm)") +
  coord_sf(expand = FALSE) +
  common_theme

# Maximum Temperature
ZNP_tmax_plot <- ggplot() +
  geom_spatraster(data = exact_ZNP_tmax) +
  geom_sf(data = ZNP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Zion National Park: Mean Annual Max Temperature", fill = "Temperature (°C)") +
  coord_sf(expand = FALSE) +
  common_theme

# Minimum Temperature
ZNP_tmin_plot <- ggplot() +
  geom_spatraster(data = exact_ZNP_tmin) +
  geom_sf(data = ZNP, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_viridis_c(option = "H", na.value = NA) +
  labs(title = "Zion National Park: Mean Annual Min Temperature", fill = "Temperature (°C)") +
  coord_sf(expand = FALSE) +
  common_theme

# Arrange ZNP plots

ZNP_Climate <- ggarrange(ZNP_precip_plot, ZNP_tmax_plot, ZNP_tmin_plot, ncol = 1)


# Combine all plots
ANP_ZNP_Climate <-  ggarrange(ANP_Climate, ZNP_Climate, ncol = 2)
print(ANP_ZNP_Climate)

# save combined plots as an image
ggsave('ANP_ZNP_Climate.png', ANP_ZNP_Climate,  width = 15, height = 15, units = "in", dpi = 300)


```



